---
title: "Final Project"
author: "Renzo Paredes"
date: "2025-04-26"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
```

# Loading Data

```{r}
Budget_vs_Spending_Data <- read_excel("C:/Users/renzo/OneDrive/Escritorio/DATA 211-01 Data Science and Visualization/Final Project/Budget_vs_Spending_Data.xlsx")
View(Budget_vs_Spending_Data)

head(Budget_vs_Spending_Data)
```

# Data Preparation

Prepare the dataset by converting the Date column to the correct type and reshaping the data for visualizations.

```{r}
Budget_vs_Spending_Data$Date <- as.Date(Budget_vs_Spending_Data$Date, format = "%m/%d/%Y")

long_data <- pivot_longer(Budget_vs_Spending_Data, 
                          cols = c(Daily_Budget, Daily_Spending),
                          names_to = "Type", 
                          values_to = "Amount")

head(long_data)
```

# Line Plot - Daily Budget vs. Daily Spending

Visualize how your daily budget and spending change over time.

```{r}
ggplot(Budget_vs_Spending_Data, aes(x = Date)) +
  geom_line(aes(y = Daily_Budget, color = "Daily Budget"), size = 1.2) +
  geom_line(aes(y = Daily_Spending, color = "Daily Spending"), size = 1.2) +
  scale_color_manual(values = c("Daily Budget" = "#1f77b4", "Daily Spending" = "#ff7f0e")) +
  labs(title = "Daily Budget vs. Daily Spending",
       x = "Date", y = "Amount (USD)", color = "Legend") +
  theme_minimal(base_size = 14)
```

# Bar Plot - Daily Comparison

Compare daily budget and daily spending side-by-side for each day.

```{r}
ggplot(long_data, aes(x = Date, y = Amount, fill = Type)) +
 geom_col(position = "dodge", width = 0.7) +
 scale_fill_manual(values = c("Daily Budget" = "#1f77b4", "Daily Spending" = "#ff7f0e")) +
 labs(title = "Daily Budget vs. Daily Spending (Side-by-Side)",
      x = "Date", y = "Amount (USD)", fill = "Type") +
 theme_light(base_size = 14) +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
 scale_x_date(date_breaks = "1 day", date_labels = "%b %d")

```


# Boxplot - Distribution and Outliers

Understand the distribution and detect potential outliers in your spending and budgeting.

```{r}
ggplot(long_data, aes(x = Type, y = Amount, fill = Type)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.size = 3) +
  scale_fill_manual(values = c("Daily Budget" = "#1f77b4", "Daily Spending" = "#ff7f0e")) +
  labs(title = "Distribution of Budget and Spending",
       x = "", y = "Amount (USD)") +
  theme_minimal(base_size = 14)
```


# Statistical Analysis: Paired t-Test

We want to test whether there is a significant difference between the mean Daily Budget and the mean Daily Spending.
$$
H_0: \mu_{\text{Daily Budget}} = \mu_{\text{Daily Spending}}
$$

$$
H_a: \mu_{\text{Daily Budget}} \ne \mu_{\text{Daily Spending}}
$$

Where:  
- \( \mu_{\text{Daily Budget}} \) = Mean of the Daily Budget  
- \( \mu_{\text{Daily Spending}} \) = Mean of the Daily Spending  

This is a **one-tailed paired t-test**.

```{r}
t_test_result <- t.test(Budget_vs_Spending_Data$Daily_Budget, 
                        Budget_vs_Spending_Data$Daily_Spending,
                        paired = TRUE,
                        alternative = "two.sided")

t_test_result
```

# Conclusion

Since the $p$-value = 0.5035 is greater than the significance level $\alpha = 0.05$, we fail to reject the null hypothesis $H_0$. At the $5\%$ significance level, we do not have enough evidence to conclude that there is a significant difference between the average Daily Budget ($\mu_{\text{budget}}$) and the average Daily Spending ($\mu_{\text{spending}}$).


